\titleNameLong{}

### 2015-12-08 FinalProject/Tex/paper/paper.tex
finished summarizing techniques

\section{Architectural Overview} \label{sec:architecture}
\subsection{Streaming Multiprocessors}
Modern unified GPUs consist of multiple streaming multiprocessors (SMs). While
there are many GPU architectures available today, the following description is
based off of NVIDIA's Maxwell architecture, which was released in 2014. These
SMs are similar to SIMD processors and typically contain the following
components:
\begin{description}
  \setlength\itemsep{1em}
  \item[A large register file] The register file is shared across all threads
  and typically contains thousands of registers, many more than are usually
  found in a CPU. NVIDIA's Maxwell architecture has 64K 32-bit registers in each
  SM.
  \item[Multiple caches] Maxwell includes a unified L1/texture cache, which used
  as a coalescing buffer for memory accesses. The buffer gathers requested data
  and sends the entire group to a warp. There are also additional caches TODO.
  \item[Shared memory] There are 96KB available per SM in Maxwell. However,
  thread blocks are limited to 48KB each. Shared memory for a thread-block is
  only accessible by threads within the same thread-block.
  \item[Multiple warp schedulers] Warp schedulers dispatch and allocate
  resources. Maxwell contains four schedulers per SM, and functional units are
  assigned to a particular scheduler. Using a power-of-two number of CUDA cores
  per SM allows for simple partitioning among schedulers. One ready warp is
  scheduled per cycle, which allows the GPU to hide the latency of stalled
  warps.
\end{description}

@@

The problems with existing implementations boil down to four characteristics:
\begin{description}
\item [High kernel density] Kernel launches are expensive, and typical
applications currently require many kernels. For example, a typical breadth-
first search requires about 3,000 device kernel launches, resulting in a high
overhead and memory footprint.
\item [Low compute intensity] While many kernels are launched using these
methods, they are usually fine-grained and do not contain high amounts of
parallelism. The average number of threads in these kernels is around 40, which
is close to the warp size and could be better served by a more fine-grained
approach.
\item [Workload similarity] The operations performed by each dynamically
launched kernel are usually similar, but contain different levels of parallelism
due to differing configurations or data.
\item [Low concurrency/scheduling efficiency] Due to the low compute intensity
of the device kernels, multiple kernels should be able to execute in parallel.
However, GPU architectures place limits on the number of different kernels that
can execute concurrently, and since each kernel has a relatively small number of
warps within it, the SMs are underutilized and struggle to hide memory
latencies.
\end{description}

@@

CAWA seeks to reduce the disparity among execution time in parallel warps using
three components:
\begin{description}
  \item[Dynamic warp criticality prediction] This predictor attempts to identify
  the critical warp that is used in the next two components. This is
  accomplished by using a per-warp criticality counter, which is based on the
  degree of instruction count disparity, a measure of how much diverging
  branches differ, as well as the number of stall cycles between two consecutive
  instructions. These two values are combined into a single criticality counter.
  \item[A greedy criticality-aware warp scheduler] This scheduler is built on
  top of the two existing warp scheduling policies, and gives more time to
  critical warps (based on the criticality counter described above). Ties are
  broken by the age of the warp, giving priority to older warps.
  \item[Criticality-aware cache prioritization] This scheme prioritizes cache
  accesses. It reserves a fixed number of L1 data cache ways for critical warps.
  The best number of ways to reserve was determined by experimental analysis and
  found to be 8 out of 16. The cache prioritization also uses a signature-based
  hit predictor to predict whether a given cache line belongs to the critical
  warp.
\end{description}

@@

\section{Reducing Memory Contention} \label{sec:memory}
\subsection{Warped Compression}
This section left as an exercise to the reader

\subsection{Managing Concurrency in Heterogeneous (CPU/GPU NoC contention)}
This section left as an exercise to the reader

\section{Compiler/SW-only design} \label{sec:software}
\subsection{Divergence Management}
This section left as an exercise to the reader

\subsection{Locality Aware Mapping}
This section left as an exercise to the reader

\subsection{(x) WeakBehaviorProgrammingAssumptions}
This section left as an exercise to the reader

>>>>

\section{Architectural Overview} \label{sec:architecture}
\subsection{Streaming Multiprocessors}
Modern unified GPUs consist of multiple streaming multiprocessors (SMs). While
there are many GPU architectures available today, the following description is
based off of NVIDIA's Maxwell architecture, which was released in 2014. These
SMs are similar to SIMD processors and typically contain the following
components:
\begin{description}
  \setlength\itemsep{0.5em}
  \item[A large register file] The register file is shared across all threads
  and typically contains thousands of registers, many more than are usually
  found in a CPU. NVIDIA's Maxwell architecture has 64K 32-bit registers in each
  SM.
  \item[Multiple caches] Maxwell includes a unified L1/texture cache, which used
  as a coalescing buffer for memory accesses. The buffer gathers requested data
  and sends the entire group to a warp. There are also additional caches TODO.
  \item[Shared memory] There are 96KB available per SM in Maxwell. However,
  thread blocks are limited to 48KB each. Shared memory for a thread-block is
  only accessible by threads within the same thread-block.
  \item[Multiple warp schedulers] Warp schedulers dispatch and allocate
  resources. Maxwell contains four schedulers per SM, and functional units are
  assigned to a particular scheduler. Using a power-of-two number of CUDA cores
  per SM allows for simple partitioning among schedulers. One ready warp is
  scheduled per cycle, which allows the GPU to hide the latency of stalled
  warps.
\end{description}

@@

The problems with existing implementations boil down to four characteristics:
\begin{description}
  \setlength\itemsep{0.5em}
  \item [High kernel density] Kernel launches are expensive, and typical
  applications currently require many kernels. For example, a typical breadth-
  first search requires about 3,000 device kernel launches, resulting in a high
  overhead and memory footprint.
  \item [Low compute intensity] While many kernels are launched using these
  methods, they are usually fine-grained and do not contain high amounts of
  parallelism. The average number of threads in these kernels is around 40, which
  is close to the warp size and could be better served by a more fine-grained
  approach.
  \item [Workload similarity] The operations performed by each dynamically
  launched kernel are usually similar, but contain different levels of parallelism
  due to differing configurations or data.
  \item [Low concurrency/scheduling efficiency] Due to the low compute intensity
  of the device kernels, multiple kernels should be able to execute in parallel.
  However, GPU architectures place limits on the number of different kernels that
  can execute concurrently, and since each kernel has a relatively small number of
  warps within it, the SMs are underutilized and struggle to hide memory
  latencies.
\end{description}

@@

CAWA seeks to reduce the disparity among execution time in parallel warps using
three components:
\begin{description}
  \setlength\itemsep{0.5em}
  \item[Dynamic warp criticality prediction] This predictor attempts to identify
  the critical warp that is used in the next two components. This is
  accomplished by using a per-warp criticality counter, which is based on the
  degree of instruction count disparity, a measure of how much diverging
  branches differ, as well as the number of stall cycles between two consecutive
  instructions. These two values are combined into a single criticality counter.
  \item[A greedy criticality-aware warp scheduler] This scheduler is built on
  top of the two existing warp scheduling policies, and gives more time to
  critical warps (based on the criticality counter described above). Ties are
  broken by the age of the warp, giving priority to older warps.
  \item[Criticality-aware cache prioritization] This scheme prioritizes cache
  accesses. It reserves a fixed number of L1 data cache ways for critical warps.
  The best number of ways to reserve was determined by experimental analysis and
  found to be 8 out of 16. The cache prioritization also uses a signature-based
  hit predictor to predict whether a given cache line belongs to the critical
  warp.
\end{description}

@@

\section{Compiler and Software Techniques} \label{sec:software}
\subsection{Divergence Management}
We have seen that complex control flow leads to divergent branches and as a
result divergent warps. Vector machines commonly handle this divergence
explicitly by using predication registers to conditionally execute statements.
Intel MIC accelerators, Cray-1 vector processors, and even AMD GPUs operate this
way \cite{DivergenceManagment}. However, NVIDIA GPUs manage divergence
implicitly. The GPU chooses one path and defers the execution of the other by
pushing the current PC and a thread mask onto the ``divergence stack.''
Reconvergence is then handled by popping off of the stack. This structure allows
for correct nesting of divergence and reconvergence. NVIDIA also allows for
software management of divergence using predicate registers and ``consensual
branches.'' These branches are only taken if all the threads within the current
warp have the same predication value. However, this functionality used very
limitedly, only when heuristics indicate a potential performance boost.

Compiler-driven approaches to divergence management exist for vector-style
architectures, but the problem is different when applied to GPUs. Traditional
compiler algorithms for predication are thread-agnostic, since they only
optimize control flow for a single thread. Data-parallel architectures such as
those in GPUs, on the other hand, must be thread-aware. One such set of
algorithms has been developed \cite{DivergenceManagment}.

These compiler algorithms remove control flow by linearizing execution paths.
This is done by first constructing thread-aware control dependence graphs
(CDGs). Predicate nodes are inserted between dependent basic blocks. The CDGs
are augmented by the compiler by adding three kinds of runtime masks: loop masks
track which threads are still executing; continue masks track which threads are
in a current iteration; and exit masks track which threads exited by specific
exit blocks. It is important to note that the resulting CDG is still valid for
processing, so later compiler passes can continue to process cyclic regions.

The algorithms also perform a few optimizations. If a branch condition is
determined to be thread-invariant, the branch is replaced with a consensual
branch and the predication is removed. This removes unnecessary work and
lightens register pressure. Branch conditions that cannot be proven thread
invariant have dynamic predicate uniformity tests inserted.

To linearize the control flow, loops are analyzed with structural analysis.
After the analysis has determined control flow, the direction of the edges in
the CFG are reversed, and the CFG is processed in a depth-first post- order
traversal. This yields a valid schedule.

Applying these algorithms results in code that is 2.7\% slower on average, with
little to no area, power, or energy implications. The main motivation for this
approach is that is allows for less hardware design complexity and verification
costs. From the programmer's point of view, there is no difference. In addition,
using divergence stack control sometimes results in nondeterministic order of
execution. Using the variance analysis in this approach allows the compiler to
make stronger guarantees at compile time.

\subsection{Locality Aware Mapping}
GPUs are good at exploiting simple parallelism, and high-level language
extensions have been developed to make writing parallel programs easier, but
nested parallelism is still quite difficult to manage. There are multiple
strategies for mapping nested patterns onto GPUs, but they all perform poorly
under certain circumstances. Three such mappings are:
\begin{description}
  \setlength\itemsep{0.5em}
  \item[1D mapping] Only parallelize the outer loop
  \item[Thread-block/thread mapping]assign each iteration of the outer pattern
  to a thread block and parallelize the inner pattern by threads within the
  block
  \item[Warp-based mapping] assign each iteration of the outer pattern to a warp
  and inner iterations to threads within a warp
\end{description}
None of these approaches is optimal across all applications. Instead, a
generalized mapping with strong support for nested mapping is needed. One such
approach has been developed \cite{LocalityAwareMappingOfNestedParallelPatterns}.

This locality-aware mapping has its own intermediate representation (IR) that
allows simple nesting of parallel patterns. These mappings contain three
parameters: dimension, a dimension for each nest level; block size, the number
of threads in a given block; and degree of parallelism (DOP), the number of
parallel computations enabled by the mapping. Analysis of these mappings is
flexible and not limited to one or two levels like the other mappings mentioned
above.

Two orthogonal categories are used to evaluate the strength of these mappings:
local vs. global and hard vs. soft. Local constraints are applied to a specific
pattern, such as having sequential accesses map to the fastest varying
dimension. Global constraints consider multiple patterns together. Hard
constraints must be satisfied to ensure correct execution, such as a limit on
the number of threads within a block. Soft constraints provide performance
hints, but can be violated if necessary. An example of a soft limit is avoiding
thread divergence. Using these constraints, mappings are ranked, and the most
efficient is chosen. Only candidates that meet hard constraints are considered,
and ties are broken in favor of higher DOP.

The most efficient mapping is then used to generate CUDA code. A few
optimizations can be applied to this generation, such as preallocating memory
before kernel launch if all loop iterations have a constant size allocation.
Another optimization uses prefetching for multiple loop iterations when loops
are imperfectly nested.

The efficiency of these mappings varies depending on the application. For simple
nested patterns, the generated code is 20\% slower than manually-optimized using
the 1D mapping. However, using the fixed 2D strategies mentioned are 1.5-9.6x
slower than the multi-dimensional mappings. On real-world applications, these
custom mappings are 1.18x-4.38x faster than running on an 8-core CPU and
4.5x-8.95x faster than using a 1D mapping.

\section{Other Research} \label{sec:memory}
\subsection{Warp Data Compression}
To increase the breadth of this survey, two additional topics are discussed that
do not fit well in the categories so far. The first is compression and
decompression of register file accesses on a warp level. GPU register files
contribute up to 15\% of the total GPU chip power, so there are significant
savings available if the register file can be made more power efficient. One
past approach has been turning off unused parts of the register file. This saves
both leakage and dynamic power. Another approach takes advantage of register use
locality with register file caches.

The technique discussed here is compressing at the warp level
\cite{WarpCompression}. Some GPU architectures already use data compression. For
example, NVIDIA's Maxwell architecture claims to improve effective bandwidth by
25\% using a compression scheme. For these applications, compression schemes
should satisfy two conditions: the data should have a high compression ratio
using the selected scheme, and the latency of compressing/decompressing should
be short. In the past, most schemes have focused on the L2 or last level cache
to avoid latency issues.

To compress the values used in warps, value similarity is key. This is the
arithmetic distance between register values. Values across threads in a warp are
typically similar, since data arrays are commonly accessed using the thread IDs
of successive threads, and the dynamic range of input data is small for many
things. Power can be saved by avoiding the repeated storing, reading, and
writing of redundant data. Distances between successive thread registers are
calculated and placed into one of four bins: zero distance, which means the two
values are the same; 128 bin, where the values differ by at most 128; 32K bin,
where they differ by at most 32K, and the random bin, which holds all values
that do not fall into any of the first three.

Keeping in mind the trade-off of compression ratio vs latency, the base-delta-
immediate (BDI) scheme is most suitable warp compression. This scheme uses a
base value, and subsequent deltas are added to the base to determine each
subsequent value. Computed values are compressed before writing to the register
file and decompressed before reading. Since most warp instructions have two
sources and one destination, it makes sense to have two decompression units and
a single compression unit to process one warp per cycle. Since modern GPU
architectures can process two warps per cycle, the number of each of these is
doubled, for a total of four decompression units and two compression units.

The results of this compression scheme are promising. The area overhead is only
0.3\% of the total SM size, and an overall 25\% power reduction (35\% dynamic,
10\% static) is achieved in the register file. While added pipeline stages for
compression and decompression add time, the total performance loss is only
0.1\%. Warp compression is clearly a promising technique for reducing power in
the massive register files of GPUs.

\subsection{Memory Contention in Heterogeneous Multiprocessing Systems}
Modern computing system architectures include both a CPU and GPU in a master-
slave configuration, in which the CPU offloads heavy parallel computations to
the GPU. In the future, there will likely be many CPUs and GPUs operating in
conjunction within a single system. These network-on-chip (NoC) configurations
are called heterogeneous multiprocessing architectures
\cite{ManagingConcurrencyInHeterogeneous}.

CPU applications tend to be more latency sensitive, while GPU applications are
more bandwidth sensitive. In such heterogeneous architectures containing both,
GPUs tend to cannibalize the performance of CPUs much more than the other way
around. The goal of the technique described here is to prevent this
cannibalization and improve the performance of both kinds of computation. This
technique assumes a tile-based mesh architecture consisting of both CPUs and
GPUs. The CPU to GPU ratio is 2:1 because a single GPU SM occupies roughly half
the space of a CPU core. The last level cache of the system is shared by CPUs
and GPUs, and they use a shared network and shared memory channels.

GPU concurrency can have a positive or negative effect on other GPU performance.
Increasing the number of concurrent warps can help performance but can also
thrash the cache or cause memory congestion. The effects of GPU concurrency on
CPU performance, however, are much more negative. Memory congestion hurts CPUs
much more than GPUs, as they do not have enough thread-level parallelism to hide
the latency.

Two schemes are presented to keep GPUs from monopolizing the resources of the
shared system. The first is CPU-Centric Concurrency Management (CM-CPU). This
scheme dynamically monitors congestion and modulates the number of active warps
on GPUs in steps of two. The second scheme is Balanced Concurrency Management
(CM-BAL), which detects if GPUs do not have enough latency tolerance, and
increases the number of active warps to hide it. The goal of this scheme is to
recover performance lost from the first scheme.

These approaches only require 16 bytes of overhead for the GPU core, 3 bytes per
memory partition, and 8 bytes for the thread-block scheduler. CM-CPU resulted in
an 11\% average GPU performance loss and a 24\% CPU performance increase. CM-BAL
resulted in a 7\% loss in GPU performance and a 7\% gain in CPU performance.
Overall speedup of the system depends on the type of application and relative
importance of CPU and GPU computation. CM-CPU ranges from 0.875x-1.25x overall
speedup, which CM-BAL yields a more balanced 1.05x-1.1x.
